Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-08-24 manuscript.Rnw
Rewrote abstract cf. NewPhyt requirements; more editing for clarity

The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs). Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical studies have not been able to justify these differences. Using a new global database on biomass partitioning from over 175 studies, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Similarly we find that PFTs differ in the amount of leaf mass per unit stem cross section, but not in the amount of leaf area per unit stem cross section. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.

@@

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key physiological attributes. While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.

In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}. Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF). \citet{poorter2012} and \citet{enquistniklas2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms. It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity. Several classic, oft-cited, studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.

@@

Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning. Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability \citep{savage2010}. Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs. This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations. Here, we use a unique, new database \citep{falster2015} (Fig.~\ref{fig:figoverview}) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. A recent compilation of plant biomass data \citep{poorter2015} speculated that differences between gymnosperms and angiosperms in distribution of biomass between leaves and stems could be related to differences in leaf mass per area. Here we can directly test this hypothesis as our database includes many observations of leaf area as well as leaf mass measured on the same plants.

@@

The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}. Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs. stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment. The two processes are different because of continuous turnover of biomass, which differs strongly between compartments. We may write \citep{McMurtrieWolf1983},

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total include records for \Sexpr{nrow(baad_all$data)} individuals. The database does not overlap with the recent compilation of \citet{poorter2015}. In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}). The database contains many more variables, for example root biomass for a much smaller subset of studies. Here we limit the analysis to patterns in aboveground biomass distribution. For each analysis, we used different subsets of the data because not all variables were measured in each study. Sample sizes by PFT are summarized in Table 1. We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

The raw data in Fig.~\ref{fig:figrawdata} show a steeper increase of woody aboveground biomass ($M_S$) with plant height than for foliage biomass ($M_F$). As a result, there was a strong effect of plant height on LMF , with the three plant functional types clearly differing in LMF across nearly the entire size range (Fig.~\ref{fig:figlmf}a). When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.~\ref{fig:figlmf}b). As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig. ~\ref{fig:figlmf}c and Fig. S2).

@@

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig.\ref{fig:fighist}). It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

@@

The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots. The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.~\ref{fig:figlmfspecies}a), explaining 30\% of the variation in $M_F/A_S$ across species. The regressions across species by PFT were broad and overlapping, with a more general relationship extending across the entire LMA axis. A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig.~\ref{fig:figlmfspecies}b). As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig.~\ref{fig:figlmf}e), but does vary close to two orders of magnitude across species. In Fig.~\ref{fig:figlmfspecies}b, the separation among PFTs in $M_F/A_S$ arises due to differences in LMA.

@@

We found that PFTs differed in LMF, and that LMF was proportional to LMA. This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT. The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) does not differ between PFTs. This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.~\ref{fig:figlmf}a). Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012,poorter2015}, but have not further separated functional types into deciduous vs. evergreen angiosperms, or provided an explanation in terms of leaf traits. We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains these differences in a consistent manner.

>>>>

% 200 word max.
(1) The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs). Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical studies have not been able to justify these differences. Here we test for differences between PFTs in partitioning of biomass between leaves and stems.
(2) We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants. The database covers the global climate space, allowing us to test for direct climate effects in addition to PFT. 
(3) The leaf mass fraction (LMF, leaf / total aboveground biomass) varied strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We found that LMF, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. We found only weak and inconsistent effects of climate on biomass partitioning.
(4) Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in GVMs.

@@

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key physiological attributes. While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.

In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}. Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012,poorter2015}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF). \citet{poorter2012} and \citet{enquistniklas2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms. It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity. Several classic, oft-cited, studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.

@@

Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning. Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability \citep{savage2010}. Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs. This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations. Here, we use a unique, new database \citep{falster2015} (Fig.~\ref{fig:figoverview}) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. A recent compilation of plant biomass data \citep{poorter2015} speculated that differences between gymnosperms and angiosperms in distribution of biomass between leaves and stems is related to differences in leaf mass per area. Here we can directly test this hypothesis as our database, unlike that of \citet{poorter2015}, includes many observations of leaf area as well as leaf mass measured on the same plants.

@@

The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}. Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs. stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment. The two processes are different because of continuous turnover of biomass, which differs strongly between compartments. We may write \citep{mcmurtriewolf1983},

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total include records for \Sexpr{nrow(baad_all$data)} individuals. The database does not overlap with the recent large compilation of \citet{poorter2015}. In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}). The database contains many more variables, for example root biomass for a much smaller subset of studies. Here we limit the analysis to patterns in aboveground biomass distribution. For each analysis, we used different subsets of the data because not all variables were measured in each study. Sample sizes by PFT are summarized in Table 1. We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

The raw data in Fig.~\ref{fig:figrawdata} show a steeper increase of woody aboveground biomass ($M_S$) with plant height compared to foliage biomass ($M_F$). As a result, LMF decreased with plant height, with the three plant functional types clearly differing in LMF across nearly the entire size range (Fig.~\ref{fig:figlmf}a). When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.~\ref{fig:figlmf}b). As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig. ~\ref{fig:figlmf}c and Fig. S2).

@@

There was considerable variation in all studied variables between species within PFTs (Fig.\ref{fig:fighist}). To understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}). The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots.

@@

The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.~\ref{fig:figlmfspecies}a), explaining 30\% of the variation in $M_F/A_S$ across species. The regressions across species by PFT were broad and overlapping, with a more general relationship extending across the entire LMA axis. A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig.~\ref{fig:figlmfspecies}b). As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig.~\ref{fig:figlmf}e), but does vary close to two orders of magnitude across species. In Fig.~\ref{fig:figlmfspecies}b, the separation among PFTs in $M_F/A_S$ arises due to differences in LMA.

@@

We found that PFTs differed in LMF, and that LMF was proportional to LMA. This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT. The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) does not differ between PFTs. This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.~\ref{fig:figlmf}a). Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012,poorter2015}, but have not been able to explain these differences in terms of leaf traits. We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains differences in LMF in a consistent manner.
